Hijacking Malaria Simulators with Probabilistic Programming

### 2019-07-04 paper/icml_ai4good/paper.tex
added tom notes

\begin{icmlauthorlist}
\icmlauthor{Bradley.~J Gram-Hansen}{equal,en}
\icmlauthor{Christian Schr\"{o}der de Witt}{equal,en}
\icmlauthor{Tom Rainforth}{st}
\icmlauthor{Phil H.S. Torr}{en}
\icmlauthor{Yee Whye Teh}{st}
\icmlauthor{At\i{}l\i{}m G\"{u}ne\c{s} Baydin}{en}
\end{icmlauthorlist}

\icmlcorrespondingauthor{Bradley.~J Gram-Hansen}{bradley@robots.ox.ac.uk}

\icmlaffiliation{en}{Department of Engineering, University of Oxford, UK}
\icmlaffiliation{st}{Department of Statistics, University of Oxford, UK}

@@

To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations \cite{who_global_2015}. 
These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lives.
For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \cite{cameron2015defining}.

@@

In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators. We achieve this by extending the work of \citep{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}. To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.  Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.

@@

Our framework already supports application to simulators written in 13 industrial programming frameworks, and is easily extensible. This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro \cite{bingham2019pyro}, is often infeasible.
%Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.

@@

This paper first gives an overview of existing malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}). Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).

@@

Two the most advanced malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
OpenMalaria is based on microsimulations of \textit{Plasmodium falciparum} in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts
Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions
Both EMOD and OpenMalaria are open source and implemented in C++.

@@

 Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
can be used to express probabilistic models and consequently perform automated inference these. Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov Chain Monte Carlo (MCMC)\cite{geyer1992practical}, Black-box Variational Inference~(VI)\cite{ranganath2014black} and amortized inference\cite{le2016inference},  can be used by non-experts in an automated fashion.

\paragraph{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS). In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators. E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}. Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.

The PPS and the simulator exchange xTensor objects \url{https://xtensor.readthedocs.io/en/latest/} through TCP or ICP using a generic FlatBuffers \url{http://google.github.io/flatbuffers/} protocol. On the PPS side, sampling is done using the deep learning framework PyTorch \cite{paszke2017automatic}.

After a variable has been sampled, it is sent to the PPS using \textit{pyprob\_cpp::sample}, alongside with its simulator source trace. This allows the PPS to construct sample trace probabilities and other summary statistics (cf. Section \ref{sec:casestudy}).

Finally, the entry point to the simulator, i.e. \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
This allows the PPS to generate rollouts from the simulator remotely. A pictorial overview of this process is depicted in Figure~\ref{fig:how}.

\paragraph{Population-based simulators} create a trace per population member, in contrast to event-based simulators that create only a single trace per \textit{forward} call~\cite{baydin2018efficient}. This means that e.g. in OpenMalaria, a standard scenario simulation rollout over the span of $3$ years  
with a population of size $n=5000$ \cite{smith2008towards} will generate about \textit{four terabytes} of raw trace data. Unlike with \citep{baydin2018efficient}, this amount of data cannot possibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient. 
In order to deal with the shortcomings of \textit{pyprob} in a population-based simulator context, we therefore extend the framework to be able to do trace analysis on the fly.

@@

\section{Case Study: Ifakara (Tanzania)}
\label{sec:casestudy}

@@

In the following, we illustrate how our method introduces a novel introspection paradigm. By extracting trace graphs from population-based simulators, policymakers can ask specific questions about properties of the model trace flow
and not just the outcomes of the model, thus providing additional interpretability to the decision making process.

To illustrate the above, we present simulation output generated from a scenario resembling local conditions in the town of Ifakara (Tanzania). Both EMOD and OpenMalaria are configured to simulate a single population node of $n=100$ and assume constant climatic conditions and no migration over the simulation period of $3$ years. Please refer to Figure \ref{fig:EIR} for the seasonal Entomological Inoculation Rate (EIR), a measure of infectivity.

\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth/2]{../plots/EIR_Ifakara.pdf}
  \vspace{-20pt}
	\caption{Seasonal Entomological Inoculation Rate (EIR) for the Ifakara scenario. Data is averaged over $30$ day periods.}
	\label{fig:EIR}
  \vspace{-10pt}
\end{figure}

@@

We can see from the addressing schemes $A_1, \ldots, A_N$ (Tables \ref{table:addresses} and \ref{table:asingleaddress}) what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria 
By having access to such diagrams users can internally evaluate and scrutinize the decisions that the simulator is making.

@@

\begin{figure*}[h!]
	\centering
	\includegraphics[width=\textwidth]{../plots/final_emod_traces__rotated.pdf}
	\includegraphics[width=\textwidth]{../plots/ewan_25_pop_100.pdf}
	\caption{Here we run two equivalent models, compare the corresponding trace paths and corresponding 
	path probabilities taken by the thousands of random variables generated internally within the simulators. \textit{\textbf{Top:}} The specified model run in EMOD. \textit{\textbf{Bottom:}} The specified model in OpenMalaria.}
	\label{fig:plotewan}
\end{figure*}

@@

In this work we have demonstrated a method 
that enables one to hijack population-based simulators, extending 
the work of \citep{baydin2018efficient}. 
We applied our method to two malaria orientated population-based simulators and 
generated a variety of trace graphs.
Finally, we have shown how our system
enables policy makers and non-experts to analyse simulator outputs in a 
way previously unavailable in the field of epidemiology.

@@

\section{Acknowledgments}
We thank Ewan Cameron of MAP at the Big Data Institute and the OpenMalaria team at the Swiss Tropical Health Institute for their time and help. BGH is supported by the EPRSC Autonomous  Intelligent Machines and Systems grant. CSW  is supported by the project Free the Drones (FreeD) under the Innovation Fund Denmark and Microsoft. YWT’s and TR’s research leading to these results has received funding from the European Research Council under the European Union’s Seventh Framework Programme (FP7/2007-2013) ERC grant agreement no. 617071. AGB and PH are supported by EPSRC/MURI grant EP/N019474/1 and AGB is also supported by Lawrence Berkeley National Lab.

>>>>

\begin{icmlauthorlist}
\icmlauthor{Bradley~J. Gram-Hansen}{equal,en}
\icmlauthor{Christian Schr\"{o}der de Witt}{equal,en}\\
\icmlauthor{Tom Rainforth}{st}
\icmlauthor{Philip H.S. Torr}{en}
\icmlauthor{Yee Whye Teh}{st}
\icmlauthor{At\i{}l\i{}m G\"{u}ne\c{s} Baydin}{en}
\end{icmlauthorlist}

\icmlcorrespondingauthor{Bradley~J. Gram-Hansen}{bradley@robots.ox.ac.uk}

\icmlaffiliation{en}{Department of Engineering Science, University of Oxford, UK}
\icmlaffiliation{st}{Department of Statistics, University of Oxford, UK}

@@

To reach the WHO's target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers are increasingly turning to evidence-based methods, thus oftentimes relying on computational simulations \cite{who_global_2015}. 
These simulations allow policymakers to infer critical information on disease dynamics and make predictions about the impacts of policies
before they are rolled out. This frequently increases the effectiveness of interventions and thus ultimately saves resources, or even lives.
For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmission \cite{cameron2015defining}.

@@

In this paper, we introduce a novel method that allows one to shed light on the inner workings of a large class of population-based stochastic simulators. We achieve this by extending the work of \citet{baydin2018efficient} by interpreting such population-based simulators as probabilistic generative models within the framework of universal probabilistic programming (UPP) \cite{le-2016-inference}. To this end, we \emph{hijack} existing simulators by overriding their internal random number generators.  Specifically, by replacing the existing low-level random number generator in a simulator with a call to a purpose-built UPP ``controller'', which can thus control, track and manipulate the stochasticity of the simulator.

@@

Our framework already supports application to simulators written in 13 general-purpose programming languages, and is easily extensible. This is crucial as, given the enormous code size and complexity, rewriting epidemiology simulators using a dedicated universal probabilistic programming language, such as Pyro \cite{bingham2019pyro}, is often infeasible.
%Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.

@@

This paper first gives an overview of existing malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}). Our approach is then demonstrated and analysed in the context of a malaria case study (Section \ref{sec:casestudy}).

@@

Two the most advanced malaria simulators, namely EMOD~\cite{bershteyn2018implementation} and OpenMalaria  \cite{smith2008towards}, have proven to be particularly valuable to policymakers.
OpenMalaria is based on microsimulations of \textit{Plasmodium falciparum} in humans and was originally developed to simulate the impacts of malaria vaccines within simple villages or districts.
Compared with OpenMalaria, EMOD is able to simulate a variety of additional drivers, including complex geographies complete with migration and a large number of policy interventions.
Both EMOD and OpenMalaria are open source and implemented in C++.

@@

 Probabilistic programming~\cite{gordon2014probabilistic,staton2016semantics,kozen1979semantics} 
can be used to express probabilistic models and consequently perform automated inference in these. Once a probabilistic model has been expressed in a probabilistic programming language, a wide range of inference techniques, such as Markov chain Monte Carlo (MCMC) \cite{geyer1992practical}, black-box variational inference~(VI) \cite{ranganath2014black} and amortized inference\cite{le2016inference}, can be used by non-experts in an automated fashion.

\paragraph{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS). In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators. E.g., given a simulator written in C++/Boost \cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}. Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.

The PPS and the simulator exchange xTensor objects\footnote{\url{https://xtensor.readthedocs.io/en/latest/}} through TCP or IPC using a generic FlatBuffers\footnote{\url{http://google.github.io/flatbuffers/}} protocol. On the PPS side, sampling is done using the deep learning framework PyTorch \cite{paszke2017automatic}.

After a variable has been sampled, using \textit{pyprob\_cpp::sample}, it is sent by the PPS to the simulator, at the same recording the simulator execution trace as a side effect. This allows the PPS to construct sample trace probabilities and other summary statistics (cf. Section \ref{sec:casestudy}).

Finally, the entry point to the simulator, i.e., \textit{main} in C++, is replaced by a special \textit{forward} call, in this case \textit{pyprob\_cpp::forward}.
This allows the PPS to generate rollouts from the simulator remotely. A pictorial overview of this process is depicted in Figure~\ref{fig:how}.

\paragraph{Population-based simulators} create a trace per population member, in contrast to event-based simulators that create only a single trace per \textit{forward} call~\cite{baydin2018efficient}. This means that, e.g., in OpenMalaria, a standard scenario simulation rollout over the span of $3$ years  
with a population of size $n=5000$ \cite{smith2008towards} will generate about \textit{four terabytes} of raw trace data. Unlike the simulator used by \citet{baydin2018efficient}, this amount of data cannot feasibly be kept in RAM, which makes \textit{a posteriori} trace analysis very inefficient. 
In order to deal with the shortcomings of \textit{pyprob} in a population-based simulator context, we therefore extend the framework to be able to do trace analysis on the fly.

@@

\section{Case Study: Ifakara, Tanzania}
\label{sec:casestudy}

@@

In the following, we illustrate how our method introduces a novel introspection paradigm. By extracting trace graphs from population-based simulators, policymakers can ask specific questions about properties of the model trace flow
and not just the outcomes of the model, thus providing additional interpretability to the decision-making process.

To illustrate the above, we present simulation output generated from a scenario resembling local conditions in the town of Ifakara, Tanzania. Both EMOD and OpenMalaria are configured to simulate a single population node of $n=100$ and assume constant climatic conditions and no migration over the simulation period of three years. Please refer to Figure \ref{fig:EIR} for the seasonal Entomological Inoculation Rate (EIR), a measure of infectivity.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth/2]{../plots/EIR_Ifakara.pdf}
  \vspace{-20pt}
	\caption{Seasonal Entomological Inoculation Rate (EIR) for the Ifakara scenario. Data is averaged over $30$ day periods.}
	\label{fig:EIR}
  \vspace{-10pt}
\end{figure}

@@

We can see from the addressing schemes $A_1, \ldots, A_N$ (Tables \ref{table:addresses} and \ref{table:asingleaddress}) what physical events are connected to each other and how outputs in EMOD are generated from a different set of procedures as compared to OpenMalaria.
By having access to such diagrams, users can internally evaluate and scrutinize the decisions that the simulator is making.

@@

\begin{figure*}[h!]
	\centering
	\includegraphics[width=\textwidth]{../plots/final_emod_traces__rotated.pdf}
	\includegraphics[width=0.89\textwidth]{../plots/ewan_25_pop_100.pdf}
	\caption{Here we run two equivalent models, compare the corresponding trace paths and corresponding 
	path probabilities taken by the thousands of random variables generated internally within the simulators. \textit{\textbf{Top:}} The specified model run in EMOD. \textit{\textbf{Bottom:}} The specified model in OpenMalaria.}
	\label{fig:plotewan}
\end{figure*}

@@

In this work we have demonstrated a method 
that enables one to hijack population-based simulators, extending 
the work of \citet{baydin2018efficient}. 
We applied our method to two malaria-orientated population-based simulators and 
generated a variety of trace graphs.
Finally, we have shown how our system
enables policy makers and non-experts to analyse simulator outputs in a 
way previously unavailable in the field of epidemiology.

@@

\section{Acknowledgments}
We thank Ewan Cameron of MAP at the Big Data Institute and the OpenMalaria team at the Swiss Tropical Health Institute for their time and help. BGH is supported by the EPRSC Autonomous  Intelligent Machines and Systems grant. CSW  is supported by the project Free the Drones (FreeD) under the Innovation Fund Denmark and Microsoft. YWT's and TR's research leading to these results has received funding from the European Research Council under the European Union's Seventh Framework Programme (FP7/2007-2013) ERC grant agreement no. 617071. AGB and PH are supported by EPSRC/MURI grant EP/N019474/1 and AGB is also supported by Lawrence Berkeley National Lab.
